Estimating Uncertainty for Tipping Points with Deep Learning

### 2022-05-12 manuscript/manuscript.Rmd
ensemble model scores + updated plots

```{r setup}
library(tidyverse)
library(patchwork)
```

@@

forecasts <- read_parquet("data/forecasts.parquet")
observations <- read_parquet("data/observations.parquet")

hopf_obs <- observations |> filter(simulation=="hopf")
stochastic_obs <- observations |> filter(simulation=="stochastic")
saddle_obs <- observations |> filter(simulation=="saddle")
```

```{r fig.cap="Forecasts of the Hopf bifurcation under each model"}
forecasts |> filter(simulation=="hopf", reps == 1, group == 1) |> 
  ggplot(aes(t, col=type, fill=type)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = hopf_obs) +
  geom_line(aes(y=mean)) +
  geom_line(aes(y=mean), data = hopf_obs) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw() + 
  ggtitle("Hopf bifurcation")
```

```{r}
forecasts |> filter(simulation=="saddle", group == 1, reps == 1) |> 
  ggplot(aes(t, col=type, fill=type)) + 
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_obs) +
  geom_line(aes(y=mean)) +
  geom_line(aes(y=mean), data = saddle_obs) +
  facet_wrap(~forecasting_model, scales = "free") +
  #facet_grid(forecasting_model ~ reps, scales = "free") +
  theme_bw() + 
  ggtitle("Saddle-Node bifurcation")
```

```{r}
forecasts |> filter(simulation=="stochastic", group == 1, reps==1) |> 
  ggplot(aes(t, col=type, fill=type)) + 
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_obs) +
  geom_line(aes(y=mean)) +
  geom_line(aes(y=mean), data = stochastic_obs) +
  facet_wrap(~forecasting_model, scales = "free") +
  #facet_grid(forecasting_model ~ reps, scales = "free") +
  theme_bw() + 
  ggtitle("Stochastic transition")
```

@@

```{r}
shift |>
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.006, show.legend = FALSE) + 
  coord_flip() +
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw()

@@

```{r}
# scores over time, averaged over replicates
over_time <- shift |> 
  group_by(t, simulation,forecasting_model, metric ) |> 
  summarise(sd = sd(score),
            score = mean(score),
            .groups = "drop") |>
  group_by(simulation,forecasting_model, metric) |> 
  mutate(t = t - min(t)) |> ungroup()

p1 <- over_time |> filter(metric == "crps") |>
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
                  shape=".", alpha=0.4, show.legend=FALSE) +
  geom_line(show.legend=FALSE) +
  facet_wrap(~simulation, scales="free", ncol=1) + 
  theme_bw()

p2 <- over_time |> filter(metric == "logs") |>
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
                  shape=".", alpha=0.4, show.legend=FALSE) +
  geom_line(show.legend=FALSE) +
  facet_wrap(~simulation, scales="free", ncol=1) + 
  scale_y_log10() +
  theme_bw()

p1 + p2
```
```{r}
over_time |> filter(metric == "logs") |>
  ggplot(aes(t, score, col = forecasting_model)) + 
  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
                  shape=".", alpha=0.4, show.legend=FALSE) +
  geom_line(show.legend=FALSE) +
 # facet_wrap(~simulation, scales="free", ncol=1) + 
  facet_grid(simulation ~ forecasting_model, scales="free") + 
  theme_bw() + scale_y_log10()
```

>>>>

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(patchwork)
library(arrow)

bound <- function(x, percentile = 0.975) {
  dplyr::nth(x, percentile * length(x), order_by = x)
}
```

@@

forecasts <- read_parquet("data/forecasts.parquet") |> filter(forecasting_model != "ml_ensemble")
observations <- read_parquet("data/observations.parquet")

hopf_obs <- observations |> mutate(variable = as.character(variable)) |> filter(iter %in% 1:15, simulation == "hopf", variable !="X")
saddle_obs <- observations |> filter(iter %in% 1:15, simulation == "saddle")
stochastic_obs <- observations |> filter(iter %in% 1:15, simulation == "stochastic")

```{r fig.cap="Forecasts of the Hopf bifurcation under each model"}
forecasts  |> 
  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1, data = hopf_obs) +
  geom_line(alpha=0.1) +
  geom_line(data = filter(hopf_obs, type=="historical")) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw()  + 
  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
  ggtitle("Hopf bifurcation")
```

forecasts |> 
  filter(iter %in% 1:15, simulation == "saddle") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1, data = saddle_obs) +
  geom_line(alpha=0.1) +
  geom_line(data = filter(saddle_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
  ggtitle("Saddle-Node bifurcation")
```

forecasts |>
  filter(iter %in% 1:15, simulation == "stochastic") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
  geom_line(alpha=0.2) +
  geom_line(data = filter(stochastic_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
  ggtitle("Stochastic transition")
```

@@

```{r}
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
  coord_flip() +
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw()

@@

```{r}
shift |> 
  group_by(simulation,forecasting_model, metric) |> 
  mutate(t = t - min(t)) |> ungroup() |> 
  filter(metric == "crps", t<100) |> 
  ggplot(aes(t, score, col = forecasting_model, fill=forecasting_model)) + 
  geom_line(aes(group=iter), alpha = 0.05, show.legend=FALSE) +
  facet_grid(simulation ~ forecasting_model, scales="free") + 
  theme_bw() 
```

```{r}
shift |> 
  group_by(simulation,forecasting_model, metric) |> 
  mutate(t = t - min(t)) |> ungroup() |> 
  filter(metric == "logs", t<100) |> 
  ggplot(aes(t, score, col = forecasting_model, fill=forecasting_model)) + 
  geom_line(aes(group=iter), alpha = 0.05, show.legend=FALSE) +
  facet_grid(simulation ~ forecasting_model, scales="free") + 
  theme_bw() + scale_y_log10()
```
